\textbf{COMP0083 Convex Optimisation Assignment}

### 2023-01-06 submission.tex
edit report
update svm plot

We are considering the optimisation problem:

\[\min_{\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty} \leq \epsilon} \frac{1}{2}\|\textbf{x}\|^2\]

@@

To compute the dual problem, we begin by reformulating the problem as:

\[\min_{\textbf{x}}  f(\textbf{x}) + g(\textbf{Ax})\]

@@

We know from the Fenchel-Rockafellar duality theory that the dual problem is formulated as:

\[\min_{\textbf{u}}  g^*(\textbf{u}) + f^*(-\textbf{A}^*\textbf{u})\]

@@

The Fenchel conjugates are:

\[g^*(\textbf{Ax}) = i^*_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty}\right) = \frac{1}{\epsilon} \| \textbf{A}\textbf{x} - \textbf{b}\|_1\]

@@

We can substitute:

\[\min_{\textbf{u}}  \frac{1}{\epsilon} \| \textbf{u} - \textbf{b}\|_1 +  \frac{1}{2}\|-\textbf{A}^*\textbf{u}\|^2\]

Simplifying, we have our dual problem:

\[\min_{\textbf{u}} \frac{1}{2}\textbf{u}^*\textbf{A}\textbf{A}^*\textbf{u} + \frac{1}{\epsilon} \| \textbf{u} - \textbf{b}\|_1\]

@@

We know that for $f(x) = \frac{1}{2}\|\textbf{x}\|^2$:

\[\textbf{0}_d \in int \big(dom(f)\big)\]

where $\textbf{0}_d = \textbf{0} \in \mathbb{R}^{d \times 1}$. Thus, knowing that $\textbf{A} \in \mathbb{R}^{n\times d}$ is a linear operator:

\[\textbf{0}_n \in int \big(\textbf{A} dom(f)\big)\]

Moreover, for $g(\textbf{u}) =  i_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{u} - \textbf{b}\|_{\infty}\right)$:

\[\textbf{0}_n \in dom(g)\]

if $\|\textbf{b}\|_{\infty} \leq \epsilon$. Therefore:

\[\textbf{0}_n \in int \big(dom(g)\big)\]

@@

The KKT conditions:

\[\textbf{x} \in \partial f^*(-\textbf{A}^*\textbf{u}) \text{ and } \textbf{Ax} \in \partial g^*(\textbf{u})\]

@@

We know that:

\[\partial f^*(-\textbf{A}^*\textbf{u}) = \partial \frac{1}{2}\|-\textbf{A}^*\textbf{u}\|^2 = -\textbf{A}^*\textbf{u}\]

Moreover:

\[\partial g^*(\textbf{u}) = \partial\frac{1}{\epsilon} \| \textbf{u} - \textbf{b}\|_1 \]

and so

\[\left(\partial\| \textbf{u} - \textbf{b}\|_1\right)_i = \begin{cases}
      \frac{1}{\epsilon}, &  \text{if } | u_i - b_i| > 0 \\
      -\frac{1}{\epsilon}, &  \text{if } | u_i - b_i| < 0 \\
      [-\frac{1}{\epsilon}, \frac{1}{\epsilon}], &  \text{if } | u_i - b| = 0 \\
   \end{cases} \\
\]

@@

Thus, our KKT conditions state that:

\[\textbf{x} = -\textbf{A}^*\textbf{u}\]

\[(\textbf{Ax})_i \in \begin{cases}
      \frac{1}{\epsilon}, &  \text{if } | u_i - b_i| > 0 \\
      -\frac{1}{\epsilon}, &  \text{if } | u_i - b_i| < 0 \\
      [-\frac{1}{\epsilon}, \frac{1}{\epsilon}], &  \text{if } | u_i - b| = 0 \\
   \end{cases} \\\]

@@

To derive a rate of convergence on the primal iterates from the applications of FISTA (Fast Iterative Shrinkage Threshold Algorithm) on the  dual problem when strong duality holds, the distance to the primal solution is bounded by the dual objective values:

\[\frac{2}{\mu} \|\textbf{x}_k - \bar{\textbf{x}}\|^2 \leq \Psi(\textbf{u}_k) - \Psi(\bar{\textbf{u}})\]

@@

We also know for FISTA that (proof in lecture):

\[\Psi(\textbf{u}_k) - \Psi(\bar{\textbf{u}}) \leq \frac{\|\textbf{u}_0 - \bar{\textbf{u}}\|^2}{2 \gamma \left(t_{k-1}}\right)^2\]

@@

Thus:

\[ \|\textbf{x}_k - \bar{\textbf{x}}\|^2 \leq \frac{\|\textbf{u}_0 - \bar{\textbf{u}}\|^2}{4 \gamma \left(t_{k-1}}\right)^2\]

Simplifying:

\[ \|\textbf{x}_k - \bar{\textbf{x}}\| \leq \frac{\|\textbf{u}_0 - \bar{\textbf{u}}\|}{2  \sqrt{\gamma}t_{k-1}}\]

Two choices for $t_k$ in FISTA are:

\begin{itemize}
\setlength{\itemindent}{2em}
\item $t_k = \frac{1+\sqrt{1+ 4 (t_{k-1})^2}}{2}$ where $\frac{1}{t_{k-1}} \leq \frac{2}{k+1}$
\item $t_k = \frac{k+a}{a}$ with $a \geq 2$ where $\frac{1}{t_{k-1}} \leq \frac{a}{k+1}$
\end{itemize}

@@

In both cases the rate of convergence of the primal iterates when applying FISTA on the dual problem:

\[ \|\textbf{x}_k - \bar{\textbf{x}}\| \leq \frac{a\|\textbf{u}_0 - \bar{\textbf{u}}\|}{\sqrt{\gamma}(k+1)}\]

@@

Our problem:

\[\min_{\textbf{x} \in \mathbb{R}^d} \frac{1}{2n} \|\textbf{Ax}-\textbf{y}\|^2 + \lambda \|x\|_1\]

@@

Equivalently, our problem:

\[\min_{\textbf{x} \in \mathbb{R}^d} \frac{1}{2n} \sum_{i=1}^n \left(\left\langle \textbf{a}^i, \textbf{x} \right\rangle - y_i \right)^2 + \lambda \|x\|_1\]

@@

The initial vector passed to the algorithms:

@@

@@

\subsection*{4.4}
Comparing the two algorithms, we can see that both are able to converge to $\alpha$ vectors producing decision boundaries that successfully separate the two training points for the two classes. However, the FISTA decision boundary is quite close to the points for both classes (i.e. the decision boundary is not as curves as for RCPGA). Thus, the quality of the solution is not as nice as the RCPGA where there is ample buffer surrounding the decision boundary for both classes. This would seem that RCPGA can produce more reasonable results, however considering the loss function plots for both algorithms, we can see that this comes at a tradeoff of computational resources/speed. FISTA is able to converge in around 50 iterations whereas RCPGA required on the order of $3\times 10^{5}$ iterations. Although each FISTA step involves two steps (to calculate $\alpha_{k+1}$ and $\nu_{k+1}$), this is still a much faster algorithm than RCGPA. Thus, depending on the practical use case, if there is a requirement on the algorithm's speed, then FISTA may be more appropriate. On the other hand, if we have more resources available, then RCGPA may be more appropriate as it seems to produce more reasonable/accurate results.

>>>>

We are considering the optimisation problem:
\[\min_{\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty} \leq \epsilon} \frac{1}{2}\|\textbf{x}\|^2\]

@@

To compute the dual problem, we begin by reformulating the problem as:
\[\min_{\textbf{x}}  f(\textbf{x}) + g(\textbf{Ax})\]

@@

We know from the Fenchel-Rockafellar duality theory that the dual problem is formulated as:
\[\min_{\textbf{u}}  g^*(\textbf{u}) + f^*(-\textbf{A}^*\textbf{u})\]

@@

The Fenchel conjugates are:
\[g^*(\textbf{Ax}) = i^*_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{A}\textbf{x} - \textbf{b}\|_{\infty}\right) = \frac{1}{\epsilon} \| \textbf{A}\textbf{x} - \textbf{b}\|_1\]

@@

We can substitute:
\[\min_{\textbf{u}}  \frac{1}{\epsilon} \| \textbf{u} - \textbf{b}\|_1 +  \frac{1}{2}\|-\textbf{A}^*\textbf{u}\|^2\]

Simplifying, we have our dual problem:
\[\min_{\textbf{u}} \frac{1}{2}\textbf{u}^*\textbf{A}\textbf{A}^*\textbf{u} + \frac{1}{\epsilon} \| \textbf{u} - \textbf{b}\|_1\]

@@

We know that for $f(x) = \frac{1}{2}\|\textbf{x}\|^2$:
\[\textbf{0}_d \in int \big(dom(f)\big)\]

where $\textbf{0}_d = \textbf{0} \in \mathbb{R}^{d \times 1}$. Thus, knowing that $\textbf{A} \in \mathbb{R}^{n\times d}$ is a linear operator:
\[\textbf{0}_n \in int \big(\textbf{A} dom(f)\big)\]

Moreover, for $g(\textbf{u}) =  i_{[0, 1]}\left(\frac{1}{\epsilon}\|\textbf{u} - \textbf{b}\|_{\infty}\right)$:
\[\textbf{0}_n \in dom(g)\]

if $\|\textbf{b}\|_{\infty} \leq \epsilon$. Therefore:
\[\textbf{0}_n \in int \big(dom(g)\big)\]

@@

The KKT conditions:
\[\textbf{x} \in \partial f^*(-\textbf{A}^*\textbf{u}) \text{ and } \textbf{Ax} \in \partial g^*(\textbf{u})\]

@@

We know that:
\[\partial f^*(-\textbf{A}^*\textbf{u}) = \partial \frac{1}{2}\|-\textbf{A}^*\textbf{u}\|^2 = -\textbf{A}^*\textbf{u}\]

Moreover:
\[\partial g^*(\textbf{u}) = \partial\frac{1}{\epsilon} \| \textbf{u} - \textbf{b}\|_1 \]

and so
\[\left(\partial\| \textbf{u} - \textbf{b}\|_1\right)_i = \begin{cases}
      \frac{1}{\epsilon}, &  \text{if } | u_i - b_i| > 0 \\
      -\frac{1}{\epsilon}, &  \text{if } | u_i - b_i| < 0 \\
      [-\frac{1}{\epsilon}, \frac{1}{\epsilon}], &  \text{if } | u_i - b| = 0 \\
   \end{cases} \\
\]

@@

Thus, our KKT conditions state that:
\[\textbf{x} = -\textbf{A}^*\textbf{u}\]

and
\[(\textbf{Ax})_i \in \begin{cases}
      \frac{1}{\epsilon}, &  \text{if } | u_i - b_i| > 0 \\
      -\frac{1}{\epsilon}, &  \text{if } | u_i - b_i| < 0 \\
      [-\frac{1}{\epsilon}, \frac{1}{\epsilon}], &  \text{if } | u_i - b| = 0 \\
   \end{cases} \\\]

@@

To derive a rate of convergence on the primal iterates from the applications of FISTA (Fast Iterative Shrinkage Threshold Algorithm) on the  dual problem when strong duality holds, the distance to the primal solution is bounded by the dual objective values:
\[\frac{2}{\mu} \|\textbf{x}_k - \bar{\textbf{x}}\|^2 \leq \Psi(\textbf{u}_k) - \Psi(\bar{\textbf{u}})\]

@@

We also know for FISTA that (proof in lecture):
\[\Psi(\textbf{u}_k) - \Psi(\bar{\textbf{u}}) \leq \frac{\|\textbf{u}_0 - \bar{\textbf{u}}\|^2}{2 \gamma \left(t_{k-1}}\right)^2\]

@@

Thus:
\[ \|\textbf{x}_k - \bar{\textbf{x}}\|^2 \leq \frac{\|\textbf{u}_0 - \bar{\textbf{u}}\|^2}{4 \gamma \left(t_{k-1}}\right)^2\]

Simplifying:
\[ \|\textbf{x}_k - \bar{\textbf{x}}\| \leq \frac{\|\textbf{u}_0 - \bar{\textbf{u}}\|}{2  \sqrt{\gamma}t_{k-1}}\]

Two choices for $t_k$ in FISTA are:
\begin{itemize}
\setlength{\itemindent}{2em}
\item $t_k = \frac{1+\sqrt{1+ 4 (t_{k-1})^2}}{2}$ where $\frac{1}{t_{k-1}} \leq \frac{2}{k+1}$
\item $t_k = \frac{k+a}{a}$ with $a \geq 2$ where $\frac{1}{t_{k-1}} \leq \frac{a}{k+1}$
\end{itemize}

@@

In both cases the rate of convergence of the primal iterates when applying FISTA on the dual problem:
\[ \|\textbf{x}_k - \bar{\textbf{x}}\| \leq \frac{a\|\textbf{u}_0 - \bar{\textbf{u}}\|}{\sqrt{\gamma}(k+1)}\]

@@

Our problem:
\[\min_{\textbf{x} \in \mathbb{R}^d} \frac{1}{2n} \|\textbf{Ax}-\textbf{y}\|^2 + \lambda \|x\|_1\]

@@

Equivalently, our problem:
\[\min_{\textbf{x} \in \mathbb{R}^d} \frac{1}{2n} \sum_{i=1}^n \left(\left\langle \textbf{a}^i, \textbf{x} \right\rangle - y_i \right)^2 + \lambda \|x\|_1\]

@@

The randomly initialised vector passed to the algorithms:

@@

We can see that both algorithms are able to recover a sparse solution, however RCPGA was able to converge in much fewer iterations compared to PSGA.

@@

\subsection*{4.4}
Comparing the two algorithms, we can see that both are able to converge to $\alpha$ vectors producing decision boundaries that are relatively successful at separating the majority of the training points for the two classes. However, the FISTA decision boundary misses a couple points for both classes with a decision boundary that is rather close to boundary points of each class. Thus, the quality of the solution is not as nice as the RCPGA where there is ample buffer surrounding the decision boundary for both classes. This would seem that RCPGA can produce better results, however considering the loss function plots for both algorithms, we can see that this comes at a tradeoff of computational resources/speed. FISTA is able to converge in around 50 iterations whereas RCPGA required on the order of $3\times 10^{5}$ iterations. Although each FISTA step involves two steps (to calculate $\alpha_{k+1}$ and $\nu_{k+1}$), this is still a much faster algorithm than RCGPA. Thus, depending on the practical use case, if there is a requirement for algorithm speed, then FISTA may be more appropriate. On the other hand, if we have more resources available, then RCGPA may be more appropriate as it seems to produce more reasonable/accurate results.
